---
title: "08_qtl"
author: "Samuel Gurr"
date: "2025-01-29"
output: html_document
---

## Objective: 

* to determine loci under selection and their heritability across generations - one method being QTL (quantitative trait locus) using packages such as lme4qtl. To run this analysis (scripts downstream from here) will will use the covariance matrix based on the relatedness (kinship) of our scallops. **In this script** will load the kinship matrix and vcf files to run qtl at each locus

### set working directory

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# SET WORKING DIRECTORY 
#knitr::opts_knit$set(root.dir = "C:/Users/samjg/Documents/Github_repositories/EAD-ASEB-Airradians_Popgen_OA/") # Sam's
knitr::opts_knit$set(root.dir = "C:/Users/samuel.gurr/Documents/Github_repositories/EAD-ASEB-Airradians_Popgen_OA/") # Sam's
```


### load libraries

```{r, include=FALSE}
library(tidyverse)
library(lme4qtl)
library(devtools)
library(vcfR)
library(lme4)
```


### Load and subset the vcf file with all SNPs

* objective: to call different vcf file subsets to run pedigree and parentage analysis

  - first lets see the calls to subset based on generation and life stage, below our master calls
  
  - **important** F0-12, F0-19, and F0-23 were sequenced but did not contribute to F1 offspring, omit
  F0-12 is column 4
  F0-19 is column 11
  F0-23 is column 16
  
```{r list all IDs in the master vcf}

# call th vcf file with all ids for the project 
All.vcf <- vcfR::read.vcfR(here::here(getwd(),
                            "RAnalysis", 
                            "Data", 
                            "Popgen", 
                            "03_prune", 
                            "out.7.phased.vcf.gz"), verbose = FALSE)

All.ListIDs    <- colnames(All.vcf@gt[,2:length(colnames(All.vcf@gt))])


```

```{r load kinsip matrix: Low v Moderate}

# load it in
LowvMod_kinship.df <- data.frame(
                               read.csv(paste0(getwd(),
                                               "/RAnalysis/Output/Popgen/pedigree/Kinship_matrix_low_v_moderate.csv")),
                                         row.names=TRUE) 
# notive the column names for some reason converted the - to ., convert back so 
# the rows are the same as columns, need this format to related back to the vcf and lifehistory data
rownames(LowvMod_kinship.df) <- gsub("\\.(F\\d+)\\.", "\\1-", rownames(LowvMod_kinship.df)) # 

# convert to matrix
LowvMod_kinship.mx           <- as.matrix(LowvMod_kinship.df)# convert to a matrix

```

## Build strata metadata 

* long custom dplyr pipe based on the id name, all correct and good to go
* assings genertaion and treatment, combine these as well for single factor if needed

```{r strata metadata}

# list ids for all individuals int he vcf file
All.ListIDs  <- colnames(All.vcf@gt[,2:length(colnames(All.vcf@gt))])

# final metadata
All.metadata <- as.data.frame(matrix(ncol = 1,
                                     nrow = length(All.ListIDs))) %>% 
                          mutate(id = All.ListIDs,
                                 type = dplyr::case_when(grepl("-B", id) ~ "broodstock", TRUE ~ 'juvenile'),
                                 gen = dplyr::case_when(grepl("F0", id) ~ "F0",
                                                             grepl("F1", id) ~ "F1",
                                                             grepl("F2", id) ~ "F2",
                                                             grepl("F3", id) ~ "F3",
                                                             TRUE ~ "F1"),
                                 treatment = dplyr::case_when(
                                        grepl("F0", id) ~ "none",
                                        grepl("pH7\\.",id) ~ "High",
                                        grepl(c("pH75\\.|.201.|.203.|.204.|.251.|.253.|.254.|.301.|.303.|.304.|.351.|.352.|.353.|.354."), id) ~
                                        "Moderate",
                                        grepl(c("pH8|.101.|.103.|.104.|.153.|.154.|.155.|.3.|.4.|.5."), id) ~ 
                                        "Low")) %>% 
                                dplyr::mutate(gen_treatment = 
                                                dplyr::case_when(gen == "F0" ~ "F0",
                                                                 gen %in% c("F1","F2","F3") ~
                                                                   paste0(gen,'_',treatment))) %>% 
                                select(-V1) 

```


# lme4qtl 


### Prepare QTL: Low v Moderate (all generations)

* prepare inputs, run checks to ensure data is balanced

(1) covariance matrix (kinship) = LowvMod_kinship.mx
  - ready to go as is, these ids need to be the ONLY ids in the following inputs #2 and 3
(2) SNP matrix
  - convert master vcf All.vcf to master genind
  - convert master genind to master matrix 
  - subset master matrix to target ids in the kinship covariance matrix
(3) metadata
  - subset master metadata to target ids in the kinship covariance matrix

```{r Prep for QTL: Low v Moderate}

# ::::::::::::::::::::::::::::::::::::::::::::::::::
# covariance matrix
# ::::::::::::::::::::::::::::::::::::::::::::::::::

# head(LowvMod_kinship.mx) # use the ids to subet the next two datasets
LowvMod_cov           <- Matrix::Matrix(LowvMod_kinship.mx, sparse = TRUE) # cov mx for lme4qtl
# note: i dont know why by matrix doesnt lke th id, converts the hyphen to a period
# i switch it back again below so that the ids match up, if we dont do this we get an error later
colnames(LowvMod_cov) <- gsub("\\.(F\\d+)\\.", "\\1-", rownames(LowvMod_cov)) # colnames gotta match


# ::::::::::::::::::::::::::::::::::::::::::::::::::
# SNP natrix
# ::::::::::::::::::::::::::::::::::::::::::::::::::

# convert master vcf to master genind
All.genind     <- All.vcf %>% vcfR::vcfR2genind()

# convert master geneind to subset matrix based on ids in the cov kinship
LowvMod_SNPs     <- tibble::rownames_to_column( # tibble wants to act on a df, move rownmaes to col
                              (as.data.frame(as.matrix(All.genind))),# conversion genind - mx - df
                              "id") %>% # call it id
                  dplyr::filter(id %in% rownames(LowvMod_kinship.mx)) %>%  # filter id for those in kinship mx
                  # lme4qtl family = binomial(link = "probit") only recognized 0,1 data 
                  # the current file before the call before contains duplicate colnames as
                  # '__.0' or '___.1' for loci, 0 being data as 2,1 and 1 being data as 0,1
                  # so filter out all columns with the string '.1' to retain only loci described with 0,1 nomenclature
                  dplyr::select(!matches("\\.0$"))

# ::::::::::::::::::::::::::::::::::::::::::::::::::
# metadata
# ::::::::::::::::::::::::::::::::::::::::::::::::::

LowvMod_metadata <- All.metadata %>% 
                          dplyr::filter(id %in% rownames(LowvMod_kinship.mx)) # filter id for those in kinship mx

# ::::::::::::::::::::::::::::::::::::::::::::::::::
# SANITY CHECK - if the three datafiles hace the same amount of data
# ::::::::::::::::::::::::::::::::::::::::::::::::::

nrow(LowvMod_metadata) == nrow(LowvMod_kinship.mx) # TRUE

nrow(LowvMod_SNPs) == nrow(LowvMod_kinship.mx) # TRUE


# ::::::::::::::::::::::::::::::::::::::::::::::::::
# merge snp with metadata as master
# ::::::::::::::::::::::::::::::::::::::::::::::::::

LowvMod_master <- merge(LowvMod_metadata, LowvMod_SNPs, by = 'id')


# ::::::::::::::::::::::::::::::::::::::::::::::::::
#  order of cov matrix matches that of the master (merged snps with metadata above)
# ::::::::::::::::::::::::::::::::::::::::::::::::::
# new 'master_reordered' for QTL 

desired_order             <- rownames(LowvMod_cov)
LowvMod_master_rownames   <- tibble::column_to_rownames(LowvMod_master, 'id')
reorder_index             <- match(desired_order, rownames(LowvMod_master_rownames))
LowvMod_master_reordered  <- LowvMod_master_rownames[reorder_index, ] %>% tibble::rownames_to_column('id')

# ::::::::::::::::::::::::::::::::::::::::::::::::::
# SANITY CHECK - does the order in fact match?
# ::::::::::::::::::::::::::::::::::::::::::::::::::

LowvMod_master_reordered$id == rownames(LowvMod_cov) #  all true
LowvMod_master_reordered$id == colnames(LowvMod_cov) #  all true

```


### Run QTL: Low v Moderate (all generations)

inputs: 
- SNP data = LowvMod_SNP.mx
- metadata = LowvMod_metadata
- covariance matrix = LowvMod_kinship.mx

```{r Run QTL: Low v Moderate}

# Before moving forward... filter the data for the specific question in mind
# below we are running qtl for all offspring in low and moderate from generations F1 - F3
# the covariance matrix does not need to include the relatedness of the F0 for this, remove them

# remove F0 from all data

# cov kinship matrix: removed all occurance of string 'F0' from rows (r) and columns (c) 
LowvMod_cov_rm_r  <- LowvMod_cov[!grepl("F0", rownames(LowvMod_cov)), , drop = FALSE]
LowvMod_cov_rm_rc <- LowvMod_cov_rm_r[, !grepl("F0", colnames(LowvMod_cov_rm_r))]

# snp data and metadata: omit columns that contain with 'F0' string 
LowvMod_master_reordered_rm <- LowvMod_master_reordered %>% 
                                  filter(!stringr::str_detect(id, "F0"))

# check, F0 removed from all data

LowvMod_master_reordered_rm$id == rownames(LowvMod_cov_rm_rc) #  all true
LowvMod_master_reordered_rm$id == colnames(LowvMod_cov_rm_rc) #  all true

# run a test on one snp

m1 <- lme4qtl::relmatGlmer(CM084264_1_65408.1 ~ # SNP DV (must be 0 or 1)
                            gen * treatment +# treatment ID
                            (1|id), # random factor id
                          LowvMod_master_reordered_rm, # data
                          relmat=list(id = LowvMod_cov_rm_rc), # covariance matrix (id relatedness)
                          family = binomial(link = "probit") # for binary SNP allele dat 
                          )
anova(m1) # does not get the p value
summary(m1) # summary stats
car::Anova(m1) # Kenward-Roger-corrected - I think this is ideal for us
```


```{r Loop run QTL (all loci): Low v Moderate}

# Call the cumulative dataframe that we will write to in the for loop below
lme4qtl.output <- data.frame() # start dataframe 
lme4qtl.loop   <- data.frame(matrix(nrow = 3, ncol = 6)) # create dataframe to save cumunalitively during for loop
colnames(lme4qtl.loop) <- c('stat',
                         'locus', 
                         'effect',
                         'df',
                         'Fvalue',
                         'pvalue') # names for comuns in the for loop

# what column in 'LowvMod_master_reordered_rm' do SNPs begin metadata listed beforehand)?
colnames(LowvMod_master_reordered_rm) # starts at column 6

# now with in information known,  run lme4qtl for each locus
for (i in 6:ncol(LowvMod_master_reordered_rm)) {
  
  LowvMod_master_reordered_rm[,6]
  

  if (length(unique(LowvMod_master_reordered_rm[,i])) == 3 | any(LowvMod_master_reordered_rm[,i] == 2) ) {
  } # nada, if the locus has 3 unique integers or if there is a 2 (not just 0,1)
  else { # run diagnostics!
    # now run diagnostics and next if else call
    # the lme4qtl model will not run if the group (pop) level
    # contains all 1 or all 0, for example if all individuals from F1_Low are 0
    
    diagnostics <- LowvMod_master_reordered_rm %>% 
                select(gen_treatment, i) %>% 
                group_by(gen_treatment) %>% #%>% 
                #summarise(total_sum = sum(i))
                dplyr::summarise(
                    n = n(),
                    sum_zeros = sum(
                       (dplyr::pick(.[[2]]) == 0)), # Count rows where value is 0
                    sum_ones = sum(
                       (dplyr::pick(.[[2]])  == 1))   # Count rows where value is 1
                  )
    
    if (any(diagnostics$sum_zeros == 0) | any(diagnostics$sum_ones == 0)) {
    } # nada, meaning that one of the group variables contained all 0 or all 1, lme4qtl wont run
    else { # diagnostics passed, run lme4qtl!
          
          # run qtl
          lme4qtl.run <- lme4qtl::relmatGlmer(LowvMod_master_reordered_rm[,i] ~
                                          #LowvMod_master_reordered_rm[,i] ~ # SNP DV (must be 0 or 1)
                                          gen * treatment +# treatment ID, both between subjsect, no slope needed
                                          (1|id), # random factor id
                                          LowvMod_master_reordered_rm, # data
                                          relmat=list(id = LowvMod_cov_rm_rc), # covariance matrix (id relatedness)
                                          family = binomial(link = "probit") # for binary SNP allele dat 
                                  )
          # get anova stats for the model
          #mod.loop <- car::Anova(lme4qtl.run) # Kenward-Roger-corrected
          mod.loop <- anova(lme4qtl.run)
          # write out to 
          lme4qtl.loop$stat[c(1:3)]       <- "lme4QTL, binomial probit"
          lme4qtl.loop$locus[c(1:3)]      <- colnames(LowvMod_master_reordered_rm[i])
          # gen
          lme4qtl.loop$effect[1]          <- "generation"
          lme4qtl.loop$df[1]              <- mod.loop$npar[1]
          lme4qtl.loop$Fvalue[1]          <- mod.loop$`F value`[1] # gen 
          lme4qtl.loop$pvalue[1]          <- stats::pf(mod.loop$`F value`[1], 
                                                        mod.loop$npar[1], 
                                                        271, 
                                                        lower.tail = FALSE)          
          # treatment
          lme4qtl.loop$effect[2]          <- "treatment"
          lme4qtl.loop$df[2]              <- mod.loop$npar[2]
          lme4qtl.loop$Fvalue[2]          <- mod.loop$`F value`[2] # gen 
          lme4qtl.loop$pvalue[2]          <- stats::pf(mod.loop$`F value`[2], 
                                                        mod.loop$npar[2], 
                                                        271, 
                                                        lower.tail = FALSE)          
          # gen * treatment
          lme4qtl.loop$effect[3]          <- "generation * treatment"
          lme4qtl.loop$df[3]              <- mod.loop$npar[3]
          lme4qtl.loop$Fvalue[3]          <- mod.loop$`F value`[3] # gen 
          lme4qtl.loop$pvalue[3]          <- stats::pf(mod.loop$`F value`[3], 
                                                        mod.loop$npar[3], 
                                                        271, 
                                                        lower.tail = FALSE)
          # cumulative output table
          lme4qtl.output <- rbind(lme4qtl.output,lme4qtl.loop) #bind to a cumulative list dataframe
          # print to keep track of it
          # print(lme4qtl.loop) # print to monitor progress - tag out to quicken loop 
    } # end of SECOND if else loop, run lme4qtl
  } # end of FIRST if else loop, diagnostics
} # end of for loop

lme4qtl.output_1 <- lme4qtl.output
unique(lme4qtl.output_1$locus)
View(lme4qtl.output_1)
```

