---
title: "08_qtl"
author: "Samuel Gurr"
date: "2025-01-29"
output: html_document
---

## Objective: 

* to determine loci under selection and their heritability across generations - one method being QTL (quantitative trait locus) using packages such as lme4qtl. To run this analysis (scripts downstream from here) will will use the covariance matrix based on the relatedness (kinship) of our scallops. **In this script** will load the kinship matrix and vcf files to run qtl at each locus

### set working directory

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# SET WORKING DIRECTORY 
knitr::opts_knit$set(root.dir = "C:/Users/samjg/Documents/Github_repositories/EAD-ASEB-Airradians_Popgen_OA/") # Sam's
#knitr::opts_knit$set(root.dir = "C:/Users/samuel.gurr/Documents/Github_repositories/EAD-ASEB-Airradians_Popgen_OA/") # Sam's
```


### load libraries

```{r, include=FALSE}
library(tidyverse)
library(lme4qtl)
library(devtools)
library(vcfR)
```


### Load and subset the vcf file with all SNPs

* objective: to call different vcf file subsets to run pedigree and parentage analysis

  - first lets see the calls to subset based on generation and life stage, below our master calls
  
  - **important** F0-12, F0-19, and F0-23 were sequenced but did not contribute to F1 offspring, omit
  F0-12 is column 4
  F0-19 is column 11
  F0-23 is column 16
  
```{r list all IDs in the master vcf}

# call th vcf file with all ids for the project 
All.vcf <- vcfR::read.vcfR(here::here(getwd(),
                            "RAnalysis", 
                            "Data", 
                            "Popgen", 
                            "03_prune", 
                            "out.7.phased.vcf.gz"), verbose = FALSE)

All.ListIDs    <- colnames(All.vcf@gt[,2:length(colnames(All.vcf@gt))])


```

```{r load kinsip matrix: Low v Moderate}

# load it in
LowvMod_kinship.df <- data.frame(
                               read.csv(paste0(getwd(),
                                               "/RAnalysis/Output/Popgen/pedigree/Kinship_matrix_low_v_moderate.csv")),
                                         row.names=TRUE) 
# notive the column names for some reason converted the - to ., convert back so 
# the rows are the same as columns, need this format to related back to the vcf and lifehistory data
rownames(LowvMod_kinship.df) <- gsub("\\.(F\\d+)\\.", "\\1-", rownames(LowvMod_kinship.df)) # 

# convert to matrix
LowvMod_kinship.mx           <- as.matrix(LowvMod_kinship.df)# convert to a matrix

```

## Build strata metadata 

* long custom dplyr pipe based on the id name, all correct and good to go
* assings genertaion and treatment, combine these as well for single factor if needed

```{r strata metadata}

# list ids for all individuals int he vcf file
All.ListIDs  <- colnames(All.vcf@gt[,2:length(colnames(All.vcf@gt))])

# final metadata
All.metadata <- as.data.frame(matrix(ncol = 1,
                                     nrow = length(All.ListIDs))) %>% 
                          mutate(id = All.ListIDs,
                                 type = dplyr::case_when(grepl("-B", id) ~ "broodstock", TRUE ~ 'juvenile'),
                                 gen = dplyr::case_when(grepl("F0", id) ~ "F0",
                                                             grepl("F1", id) ~ "F1",
                                                             grepl("F2", id) ~ "F2",
                                                             grepl("F3", id) ~ "F3",
                                                             TRUE ~ "F1"),
                                 treatment = dplyr::case_when(
                                        grepl("F0", id) ~ "none",
                                        grepl("pH7\\.",id) ~ "High",
                                        grepl(c("pH75\\.|.201.|.203.|.204.|.251.|.253.|.254.|.301.|.303.|.304.|.351.|.352.|.353.|.354."), id) ~
                                        "Moderate",
                                        grepl(c("pH8|.101.|.103.|.104.|.153.|.154.|.155.|.3.|.4.|.5."), id) ~ 
                                        "Low")) %>% 
                                dplyr::mutate(gen_treatment = 
                                                dplyr::case_when(gen == "F0" ~ "F0",
                                                                 gen %in% c("F1","F2","F3") ~
                                                                   paste0(gen,'_',treatment))) %>% 
                                select(-V1) 

```


# lme4qtl 


### Prepare QTL: Low v Moderate (all generations)

* prepare inputs, run checks to ensure data is balanced

(1) covariance matrix (kinship) = LowvMod_kinship.mx
  - ready to go as is, these ids need to be the ONLY ids in the following inputs #2 and 3
(2) SNP matrix
  - convert master vcf All.vcf to master genind
  - convert master genind to master matrix 
  - subset master matrix to target ids in the kinship covariance matrix
(3) metadata
  - subset master metadata to target ids in the kinship covariance matrix

```{r Prep for QTL: Low v Moderate}

# ::::::::::::::::::::::::::::::::::::::::::::::::::
# covariance matrix
# ::::::::::::::::::::::::::::::::::::::::::::::::::

head(LowvMod_kinship.mx) # use the ids to subet the next two datasets
LowvMod_cov           <- Matrix::Matrix(LowvMod_kinship.mx, sparse = TRUE) # cov mx for lme4qtl
# note: i dont know why by matrix doesnt lke th id, converts the hyphen to a period
# i switch it back again below so that the ids match up, if we dont do this we get an error later
colnames(LowvMod_cov) <- gsub("\\.(F\\d+)\\.", "\\1-", rownames(LowvMod_cov)) # colnames gotta match


# ::::::::::::::::::::::::::::::::::::::::::::::::::
# SNP natrix
# ::::::::::::::::::::::::::::::::::::::::::::::::::

# convert master vcf to master genind
All.genind     <- All.vcf %>% vcfR::vcfR2genind()

# convert master geneind to subset matrix based on ids in the cov kinship
LowvMod_SNPs     <- tibble::rownames_to_column( # tibble wants to act on a df, move rownmaes to col
                              (as.data.frame(as.matrix(All.genind))),# conversion genind - mx - df
                              "id") %>% # call it id
                  dplyr::filter(id %in% rownames(LowvMod_kinship.mx)) # filter id for those in kinship mx

# ::::::::::::::::::::::::::::::::::::::::::::::::::
# metadata
# ::::::::::::::::::::::::::::::::::::::::::::::::::

LowvMod_metadata <- All.metadata %>% 
                          dplyr::filter(id %in% rownames(LowvMod_kinship.mx)) # filter id for those in kinship mx

# ::::::::::::::::::::::::::::::::::::::::::::::::::
# SANITY CHECKS! 
# ::::::::::::::::::::::::::::::::::::::::::::::::::

nrow(LowvMod_metadata) == nrow(LowvMod_kinship.mx) # TRUE

nrow(LowvMod_SNPs) == nrow(LowvMod_kinship.mx) # TRUE


# ::::::::::::::::::::::::::::::::::::::::::::::::::
# merge snp with metadata as master
# ::::::::::::::::::::::::::::::::::::::::::::::::::

LowvMod_master <- merge(LowvMod_metadata, LowvMod_SNPs, by = 'id')

```


### Run QTL: Low v Moderate (all generations)

inputs: 
- SNP data = LowvMod_SNP.mx
- metadata = LowvMod_metadata
- covariance matrix = LowvMod_kinship.mx

```{r Run QTL: Low v Moderate}

# reorder so the cov matrix matches the master 

desired_order             <- rownames(LowvMod_cov)
LowvMod_master_rownames   <- tibble::column_to_rownames(LowvMod_master, 'id')
reorder_index             <- match(desired_order, rownames(LowvMod_master_rownames))
LowvMod_master_reordered  <- LowvMod_master_rownames[reorder_index, ] %>% tibble::rownames_to_column('id')

# check 

LowvMod_master_reordered$id == rownames(LowvMod_cov) #  all true
LowvMod_master_reordered$id == colnames(LowvMod_cov) #  all true

# remove F0 from all data

LowvMod_cov_rm_r  <- LowvMod_cov[!grepl("F0", rownames(LowvMod_cov)), , drop = FALSE]
LowvMod_cov_rm_rc <- LowvMod_cov_rm_r[, !grepl("F0", colnames(LowvMod_cov_rm_r))]

LowvMod_master_reordered_rm <- LowvMod_master_reordered %>% 
                                  filter(!stringr::str_detect(id, "F0"))

# check, F0 removed from all data

LowvMod_master_reordered_rm$id == rownames(LowvMod_cov_rm_rc) #  all true
LowvMod_master_reordered_rm$id == colnames(LowvMod_cov_rm_rc) #  all true

# run a test on one snp

m1 <- lme4qtl::relmatGlmer(CM084264_1_65408.1 ~ # SNP DV (must be 0 or 1)
                            gen * treatment +# treatment ID
                            (1|id), # random factor id
                          LowvMod_master_reordered_rm, # data
                          relmat=list(id = LowvMod_cov_rm_rc), # covariance matrix (id relatedness)
                          family = binomial(link = "probit") # for binary SNP allele dat 
                          )
anova(m1) # get the p value
summary(m1)

car::Anova(m1) # Kenward-Roger-corrected
afex::mixed(m1)
drop1(m1)
```

```{r}

data(dat40, package = "lme4qtl")

dim(dat40)
kin2[1:5, 1:5]

View(F0BF1O_kinship.mx.2)
# subset the metadata for ids in the covariance matrix
F0BF1O.metadata      <- All.metadata %>% dplyr::filter(id %in% rownames(F0BF1O_kinship.mx))
F0BF1O_kinship.mx.2  <- subset(F0BF1O_kinship.mx, 
                               colnames(F0BF1O_kinship.mx) %in% F0BF1O.metadata$id, 
                               rownames(F0BF1O_kinship.mx) %in% F0BF1O.metadata$id)
F0BF1O.metadata.2    <- F0BF1O.metadata %>% dplyr::filter(id %in% rownames(F0BF1O_kinship.mx.2))
rownames(F0BF1O.GenoM) <- gsub("\\-", ".", rownames(F0BF1O.GenoM))
F0BF1O.GenoM.2  <- as.data.frame(
  subset(F0BF1O.GenoM, rownames(F0BF1O.GenoM) %in% rownames(F0BF1O_kinship.mx.2))
) %>% tibble::rownames_to_column(., "id")

# sanity check so covarinace matrix matches the metadata
length(rownames(F0BF1O.GenoM.2)) == nrow(F0BF1O.metadata.2)
length(rownames(F0BF1O.GenoM.2)) == length(rownames(F0BF1O_kinship.mx.2))


m <- Matrix(F0BF1O_kinship.mx.2, sparse = TRUE)
dim(m )
# run lme4qtl

F0BF1O_master <- merge(F0BF1O.metadata.2, F0BF1O.GenoM.2, by = 'id')
dim(F0BF1O_master) # has the have the same number of rows and columns!

m1 <- lme4qtl::relmatLmer(CM084264_1_39013.0 ~ gen_treatment + (1|id), 
                    F0BF1O_master, 
                    relmat=list(id = m))


m1 <- lme4qtl::relmatGlmer(CM084264_1_39013.1 ~ 
                             # gen + 
                             # treatment +
                             gen_treatment +
                             (1|id), 
                    F0BF1O_master, 
                    relmat=list(id = m),
                    family = binomial(link = "probit"))

anova(m1) # get the p value

```


