---
title: "waples_test"
author: "Samuel Gurr"
date: "2025-04-03"
output: html_document
---

## Objective: 

* at this point in our pipeline (pedigree -< kinship -> binary QTL -> adj p -> annotate sig loci -> plot) we have reduced SNPs down to those putatively selected in our treatment groups. Now we want to know, how does the minor allele frequency behave across time? To this end, we 
need to convert the vcf file data to MAF and we can do this in R with the package 'gwscaR'. With MAF by loci, we can integrate the strata
populations (F1 juv, F1 brood, F2 juv.. etc.) to plot all loci and *importantly* truncate loci to those affected by treatment (via lme4QTL)
to see how the MAF behaves across lifespan and generational timescales. 

### set working directory

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# SET WORKING DIRECTORY 
knitr::opts_knit$set(root.dir = "C:/Users/samjg/Documents/Github_repositories/EAD-ASEB-Airradians_Popgen_OA/") # Sam's
#knitr::opts_knit$set(root.dir = "C:/Users/samuel.gurr/Documents/Github_repositories/EAD-ASEB-Airradians_Popgen_OA/") # Sam's
```


### load libraries

```{r, include=FALSE}
library(tidyverse)
library(ggplot2)
library(tidyverse)
library(vcfR)
library(gwscaR) #remotes::install_github("spflanagan/gwscaR")
```


### Convert vcf to MAF 

* Objective: 

```{r load master vcf file}

# load master vcf
All.vcf <- vcfR::read.vcfR(here::here(getwd(),
                            "RAnalysis", 
                            "Data", 
                            "Popgen", 
                            "03_prune", 
                            "out.7.phased.vcf.gz"), verbose = FALSE)

# filter out F0, retain all other individuals!
F1F2F3_All.vcf <- All.vcf[,c(1,27:392)]  # QTL run with all progeny

# filter out F0 and High treatment, reatain low v Mod # QTL run with just Low and Moderate progeny
F1F2F3_LowvMod.vcf <- All.vcf[,c(1,27:139, #F1
                              153:177, 198:250, # F2
                              295:392)] # F3

```


```{r custom code of vcf filenames to aquire pop metadata}

# list ids for all individuals int he vcf file
All.ListIDs  <- colnames(All.vcf@gt[,2:length(colnames(All.vcf@gt))])

# master metadata
All.metadata <- as.data.frame(matrix(ncol = 1,
                                     nrow = length(All.ListIDs))) %>% 
                          mutate(id = All.ListIDs,
                                 type = dplyr::case_when(grepl("-B", id) ~ "broodstock", TRUE ~ 'juvenile'),
                                 gen = dplyr::case_when(grepl("F0", id) ~ "F0",
                                                             grepl("F1", id) ~ "F1",
                                                             grepl("F2", id) ~ "F2",
                                                             grepl("F3", id) ~ "F3",
                                                             TRUE ~ "F1"),
                                 treatment = dplyr::case_when(
                                        grepl("F0", id) ~ "none",
                                        grepl("pH7\\.",id) ~ "High",
                                        grepl(c("pH75\\.|.201.|.203.|.204.|.251.|.253.|.254.|.301.|.303.|.304.|.351.|.352.|.353.|.354."), id) ~
                                        "Moderate",
                                        grepl(c("pH8|.101.|.103.|.104.|.153.|.154.|.155.|.3.|.4.|.5."), id) ~ 
                                        "Low")) %>% 
                                dplyr::mutate(gen_treatment = 
                                                dplyr::case_when(gen == "F0" ~ "F0",
                                                                 gen %in% c("F1","F2","F3") ~ paste0(gen,'_',treatment))) %>% 
                                select(-V1) 

```



```{r subset metadata by target vcf files}

All.metadata # here for the All.vcf

F1F2F3_All.metadata     <- All.metadata %>% dplyr::filter(id %in%
                                                            colnames(F1F2F3_All.vcf@gt[,2:length(colnames(F1F2F3_All.vcf@gt))]))

F1F2F3_LowvMod.metadata  <- All.metadata %>% dplyr::filter(id %in%
                                                             colnames(F1F2F3_LowvMod.vcf@gt[,2:length(colnames(F1F2F3_LowvMod.vcf@gt))]))
```


### Calculate Allele Frequencies by Population


```{r extract genotype information}

All.gt            <- extract.gt(All.vcf, element = "GT")


F1F2F3_All.gt     <- extract.gt(F1F2F3_All.vcf, element = "GT")


F1F2F3_LowvMod.gt <- extract.gt(F1F2F3_LowvMod.vcf, element = "GT")

```

```{r define populations}

All.strata            <- All.metadata %>% select(gen_treatment,id) # as F0 F1_Low, F1_Moderate, and so on

F1F2F3_All.strata     <- F1F2F3_All.metadata %>% select(gen_treatment,id) # as F0 F1_Low, F1_Moderate, and so on

F1F2F3_LowvMod.strata  <- F1F2F3_LowvMod.metadata %>% select(gen_treatment,id) # as F0 F1_Low, F1_Moderate, and so on

```

```{r function to compute allele frequencies}

calculate_af <- function(gt.matrix, ids) {
                        pop_gt    <- gt.matrix[, ids] # subset for each strate group
                        alt_count <- apply(pop_gt, 1, function(x)  {
                                      alt_count <- sum(as.numeric(x != "0|0"))
                                      })
                        ref_count <- apply(pop_gt, 1, function(x)  {
                                      sum(as.numeric(x == "0|0"))
                                      })
                        afs        <- cbind(data.frame(alt_count), data.frame(ref_count))
                        afs$locus  <- rownames(pop_gt)
                        afs$total_alleles <- afs$ref_count + afs$alt_count
                        afs$af_ref <- afs$ref_count / afs$total_alleles
                        afs$af_alt <- afs$alt_count / afs$total_alleles
                        return(c(afs))
}



```



```{r compute allele frequencies for each pop}
# Loop through populations iwth lapplay outputing new subset


All.afs <- lapply(unique(All.strata$gen_treatment), function(pop) {
  calculate_af(All.gt, All.strata)
})
View(All.afs)





F1F2F3_All.loop <- data.frame(unique(F1F2F3_All.strata))
F1F2F3_All.afs  <- data.frame()
for (i in 1:nrow(F1F2F3_All.loop)) {
  
  pop            = F1F2F3_All.loop$gen_treatment[i]
  ids            = (F1F2F3_All.strata %>% filter(gen_treatment %in% pop))$id 
  loop.afs       = calculate_af(F1F2F3_All.gt, ids)
  F1F2F3_All.afs = rbind(F1F2F3_All.afs, 
                          (data.frame(loop.afs) %>% mutate(Pop = pop))
                          )

}




F1F2F3_LowvMod.loop <- data.frame(unique(F1F2F3_LowvMod.strata))
F1F2F3_LowvMod.afs  <- data.frame()
for (i in 1:nrow(F1F2F3_LowvMod.loop)) {
  
  pop            = F1F2F3_LowvMod.loop$gen_treatment[i]
  ids            = (F1F2F3_LowvMod.strata %>% filter(gen_treatment %in% pop))$id 
  loop.afs       = calculate_af(F1F2F3_LowvMod.gt, ids)
  F1F2F3_LowvMod.afs = rbind(F1F2F3_LowvMod.afs, 
                          (data.frame(loop.afs) %>% mutate(Pop = pop))
                          )

}

```


# Load in the significant loci from QTL

* load in the significant SNPs from lme4qtl with stepwise bonferroni correctio, review the qtl script for details..

```{r load significant qtl loci}


# read lme4qtl results
F1F2F3_LowvMod.qtl_interaction <- read.csv("RAnalysis/Output/Popgen/qtl/LowvMod/results/LowvMod_QTL_bonferroni_genXtreatment.csv", sep = ',') %>%
                                          select (-X)
F1F2F3_LowvMod.qtl_generation  <- read.csv("RAnalysis/Output/Popgen/qtl/LowvMod/results/LowvMod_QTL_bonferroni_gen_only.csv", sep = ',') %>%  
                                          select (-X)
F1F2F3_LowvMod.qtl_treatment   <- read.csv("RAnalysis/Output/Popgen/qtl/LowvMod/results/LowvMod_QTL_bonferroni_treatment_only.csv", sep = ',') %>%
                                          select (-X)


# read lme4qtl results
F1F2F3_All.qtl_interaction <- read.csv("RAnalysis/Output/Popgen/qtl/All/results/All_QTL_bonferroni_genXtreatment.csv", sep = ',') %>%  
                                          select (-X)
F1F2F3_All.qtl_generation  <- read.csv("RAnalysis/Output/Popgen/qtl/All/results/All_QTL_bonferroni_gen_only.csv", sep = ',') %>%  
                                          select (-X)
F1F2F3_All.qtl_treatment   <- read.csv("RAnalysis/Output/Popgen/qtl/All/results/All_QTL_bonferroni_treatment_only.csv", sep = ',') %>%  
                                          select (-X)

```

### Visualize all and target loci 

  * figure for all SNPs from .afs file 
  
  * truncate .afs for target loci from qtl, visualize alongside the figure in bullet #1

```{r F1F2F3_LowvMod treatment effect}

# allelle freq master file == F1F2F3_LowvMod.afs_master
# qtl targe loci           == F1F2F3_LowvMod.qtl_treatment

F1F2F3_LowvMod.afs_qtl_treatment <- F1F2F3_LowvMod.afs %>% 
                                              dplyr::filter(locus %in% 
                                                              gsub("^(CM\\d+)_1_(\\d+)\\.(\\d+)$", "\\1.\\3_\\2", 
                                                                   F1F2F3_LowvMod.qtl_treatment$locus)) %>% 
                                    tidyr::pivot_longer(cols = c(af_ref, af_alt), 
                                                        names_to = "AlleleType",
                                                        values_to = "Frequency") %>% 
                                    dplyr::mutate(Treatment = gsub('.*_', '',Pop),
                                                  Generation = gsub('_.*', '',Pop))
# sanity check
length(unique(F1F2F3_LowvMod.afs_qtl_treatment$locus)) == nrow(F1F2F3_LowvMod.qtl_treatment)

# calculate the meanand standard error
F1F2F3_LowvMod.afs_qtl_treatmentMEANSE <-  Rmisc::summarySE(F1F2F3_LowvMod.afs_qtl_treatment, 
                                                 measurevar = ('Frequency'), 
                                                 groupvars = c('Pop','Generation','Treatment','AlleleType'))

# plot 
scatter <- ggplot(data = (F1F2F3_LowvMod.afs_qtl_treatment %>% 
                            # filter(locus %in% 'CM084266.1_65580309') %>% 
                            filter(AlleleType %in% 'af_alt')), 
       aes(y = as.numeric(Frequency), x = Generation, group = locus)) +
  geom_line(aes(color = Treatment)) +
  geom_point(aes(color = Treatment,alpha = 0.5))+
  scale_color_manual(values = c("#009E73","#E69F00")) + 
  ggtitle("Scatter MAF") +
  facet_wrap(~Treatment) + 
  theme_classic() +
  theme(legend.position = "none") 
scatter

StackedBar <- ggplot(F1F2F3_LowvMod.afs_qtl_treatmentMEANSE, 
                     aes(fill = forcats::fct_relevel(AlleleType, "af_ref", "af_alt"), 
                         y = as.numeric(Frequency), 
                         x = Generation)) +
  geom_bar(position = "stack", stat = "identity", alpha = 0.5) +
  scale_fill_manual(values = c("grey70", "grey20")) +
  ggtitle("Stacked Bar MAF v Ref") +
  facet_wrap(~Treatment) + 
  theme_classic() +
  theme(legend.position = "none") 

ggpubr::ggarrange(scatter, StackedBar, nrow = 1)
```



```{r F1F2F3_All treatment effect}

# allelle freq master file == F1F2F3_LowvMod.afs_master
# qtl targe loci           == F1F2F3_LowvMod.qtl_treatment

F1F2F3_All.afs_qtl_treatment <- F1F2F3_All.afs %>% 
                                              dplyr::filter(locus %in% 
                                                              gsub("^(CM\\d+)_1_(\\d+)\\.(\\d+)$", "\\1.\\3_\\2", 
                                                                   F1F2F3_All.qtl_treatment$locus)) %>% 
                                    tidyr::pivot_longer(cols = c(af_ref, af_alt), 
                                                        names_to = "AlleleType",
                                                        values_to = "Frequency") %>% 
                                    dplyr::mutate(Treatment = gsub('.*_', '',Pop),
                                                  Generation = gsub('_.*', '',Pop))
# sanity check
length(unique(F1F2F3_All.afs_qtl_treatment$locus)) == nrow(F1F2F3_All.qtl_treatment)

# calculate the meanand standard error
F1F2F3_All.afs_qtl_treatmentMEANSE <-  Rmisc::summarySE(F1F2F3_All.afs_qtl_treatment, 
                                                 measurevar = ('Frequency'), 
                                                 groupvars = c('Pop','Generation','Treatment','AlleleType'))


# plot 

scatter <- ggplot(data = (F1F2F3_All.afs_qtl_treatment %>% 
                            # filter(locus %in% 'CM084264.1_64874049') %>% 
                            filter(AlleleType %in% 'af_alt') %>% 
                            mutate(Treatment = forcats::fct_relevel(Treatment, "Low", "Moderate", "High"))), 
       aes(y = as.numeric(Frequency), x = Generation, group = locus)) +
  geom_line(aes(color = Treatment)) +
  geom_point(aes(color = Treatment,alpha = 0.5))+
  scale_color_manual(values = c("#009E73","#E69F00", "#CC79A7")) + 
  ggtitle("Scatter MAF") +
  facet_wrap(~Treatment) + 
  theme_classic() +
  theme(legend.position = "none") 
scatter

StackedBar <- ggplot(data = (F1F2F3_All.afs_qtl_treatmentMEANSE %>% 
                              mutate(Treatment = forcats::fct_relevel(Treatment, "Low", "Moderate", "High"))), 
                     aes(fill = forcats::fct_relevel(AlleleType, "af_ref", "af_alt"), 
                         y = as.numeric(Frequency), 
                         x = Generation)) +
  geom_bar(position = "stack", stat = "identity", alpha = 0.5) +
  scale_fill_manual(values = c("grey70", "grey20")) +
  ggtitle("Stacked Bar MAF v Ref") +
  facet_wrap(~Treatment) + 
  theme_classic() +
  theme(legend.position = "none") 

ggpubr::ggarrange(scatter, StackedBar, nrow = 1)
 
 
```


```{r waples test}

# Function to perform Waples' test for allele frequency changes
waples_test <- function(p0, pt, S0, St, t) {
  F_observed <- (pt - p0)^2 / (p0 * (1 - p0))
  F_corrected <- F_observed - (1/(2*S0) + 1/(2*St))
  Ne_estimate <- t / (2 * F_corrected)  # Effective population size estimate[3]
  
  # Chi-squared test adjusted for drift
  chi_stat <- sum((pt - p0)^2 / (p0 * (1 - p0) * (1/(2*S0) + 1/(2*St))))
  p_value <- pchisq(chi_stat, df = length(p0) - 1, lower.tail = FALSE)
  
  return(list(Ne = Ne_estimate, chi_sq = chi_stat, p_value = p_value))
}


```


```{r example usage}
# Example usage:
p0 <- c(0.8, 0.2)  # Initial allele frequencies (e.g., A1 and A2)
pt <- c(0.75, 0.25)  # Frequencies after t generations
S0 <- 100  # Initial sample size
St <- 100  # Post-drift sample size
t <- 5  # Generations between samples

result <- waples_test(p0, pt, S0, St, t)
print(result)
```


```{r}


F1F2F3_All.afs_qtl_lociMEANSE <-  Rmisc::summarySE(F1F2F3_All.afs_qtl_treatment, 
                                                 measurevar = ('Frequency'), 
                                                 groupvars = c('Pop','Generation','Treatment','AlleleType', 'locus'))


F1Mod_alt <- (F1F2F3_All.afs_qtl_lociMEANSE %>% filter(AlleleType %in% 'af_alt' & Pop %in% 'F1_Moderate'))$Frequency # 0.3604743
F3Mod_alt <- (F1F2F3_All.afs_qtl_lociMEANSE %>% filter(AlleleType %in% 'af_alt' & Pop %in% 'F3_Moderate'))$Frequency # 0.4055013
# Example usage:
p0 <- F1Mod_alt  # Initial allele frequencies (e.g., A1 and A2)
pt <- F3Mod_alt  # Frequencies after t generations
S0 <- 55  # Initial sample size
St <- 55  # Post-drift sample size
t <- 2  # Generations between samples

result <- waples_test(p0, pt, S0, St, t)
print(result)
```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


